---
title: "index"
author: "Taavi Päll"
date: "27 9 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(stringr)
```


## Download dataset
```{r}
url <- "https://raw.githubusercontent.com/rstats-tartu/datasets/master/cancer_newcases_PK10.json"
dir.create("data")
download.file(url, "data/incidence.json")
```

Install boulder library to import json.
```{r}
# install.packages("devtools")
# devtools::install_github("tpall/boulder")
```

## Import dataset
```{r}
library(boulder)
incidence <- json_to_df("data/incidence.json")
```

Load tidyverse libraries:
```{r}
library(tidyverse)
library(stringi)
library(stringr) # this is what you are going to need most in future
```


For better overview lets filter out "Specified site" values that are longer than 30 characters.
```{r}
filter(incidence, str_length(`Specified site`) < 30)
```

Convert Year to numeric
```{r}
incidence <- incidence %>% mutate_at("Year", parse_number)
```

Specified site and sex to characters
```{r}
?mutate_at
incidence <- incidence %>% mutate_at(c("Specified site","Sex"), parse_character)
incidence
```


```{r}
label <- unique(incidence$label)
data_source <- unique(incidence$source)
updated <- unique(incidence$updated)
label
data_source
updated
```

```{r}
incidence <- incidence %>% 
  mutate(Site = stri_trim_left(`Specified site`, "[[:alpha:]]")) %>% 
  select(Year, Site, Sex, everything(), -`Specified site`, -label, -source, -updated)
```

```{r}
incidence
```

It's not what we want...
```{r}
gather(incidence)
```

Convert incidence from wide to long: 
```{r}
incidence_long <- gather(incidence, key = age, value = cases, matches("[[:digit:]]"))
# gather(incidence, key = age, value = cases, -Year, -Sex, -Site)
incidence_long
```

Let's filter out NA-s.
```{r}
incidence_long <- drop_na(incidence_long)
```

That's all with data wrangling for now!

## Cancer trend in Estonia

Plot number of all cancer cases (Men + Women) per Year:
```{r}
total_cases_per_year <- incidence_long %>% 
  filter(Site == "All malignant neoplasms (C00-C97)") %>% 
  group_by(Year) %>% 
  summarise(total_cases = sum(cases))
total_cases_per_year
```

```{r}
p <- total_cases_per_year %>% 
  ggplot(aes(Year, total_cases)) +
  geom_point() +
  geom_line() +
  ylab(label) +
  ggtitle("All malignant neoplasms (C00-C97)",
          subtitle = paste("Source:", data_source)) +
  expand_limits(y = 0) +
  geom_smooth(method = "lm", se = FALSE)
p
```
Ok, trend is increasing, but how much?

Percent increase of cancer cases from 2000 to 2014? 
```{r}
range(total_cases_per_year$total_cases)
```

Please compare number of cases in male and females:
```{r}
total_cases_per_year_sex <- incidence_long %>% 
  filter(Site == "All malignant neoplasms (C00-C97)") %>% 
  group_by(Year, Sex) %>% 
  summarise(total_cases = sum(cases))
```

here we replace data and aes in previous plot called "p":
```{r}
p %+% total_cases_per_year_sex + aes(Year, total_cases, color = Sex)
```

Perhaps we want to completely rewrite our plot and throw out linear model lines and expand limits option:
```{r}
total_cases_per_year_sex %>% 
  ggplot(aes(Year, total_cases, color = Sex)) +
  geom_point() +
  geom_line() +
  ylab(label) +
  ggtitle("All malignant neoplasms (C00-C97)",
          subtitle = paste("Source:", data_source))
```

Lets get population size data for year, sex and per age:

```{r}
download.file("https://raw.githubusercontent.com/rstats-tartu/datasets/master/mean_annual_population.csv",
              "data/mean_annual_population.csv")
```

Import population dataset into R:
```{r}
pop <- read_csv("data/mean_annual_population.csv")
pop
```

Add age groups 0 and 1-4 into age group 0-4:
```{r}
pop_long <- pop %>% mutate("0-4" = `0` + `1-4`) %>% 
  select(Sex, Year, `0-4`, everything()) %>% 
  select(-`0`, -`1-4`) %>% 
  gather("age", "ave_pop", matches("[0-9]"))
pop_long
```


```{r}
incidence_long <- left_join(incidence_long, pop_long)
```

Cancer incidence is give new cases per 100'000.
```{r}
incidence_year_sex <- incidence_long %>% 
  group_by(Site, Year, Sex) %>% 
  summarise(cases = sum(cases),
            ave_pop = sum(ave_pop)) %>% 
  mutate(incidence = cases*100000 / ave_pop)
incidence_year_sex
```


```{r}
incidence_year_sex %>% filter(Site == "All malignant neoplasms (C00-C97)")
```


```{r}
incidence_year_sex %>% 
  filter(Site == "All malignant neoplasms (C00-C97)") %>% 
  ggplot(aes(Year, incidence, color = Sex)) +
  geom_point() +
  geom_line() +
  ylab("Cancer incidence per 100'000") +
  ggtitle("All malignant neoplasms (C00-C97)",
          subtitle = paste("Source:", data_source))
```

## What are the most frequent cancer sites

We have 73 cancer sites in our dataset:
```{r}
incidence_year_sex$Site %>% unique() %>% length()
```

## What are the most frequent cancer sites?

Let's have a look at Top10 most frequent cancers.
Most frequent cancers are different in men and women.

First, we try to remove some summary variables from table.
"boulder" package has *adhoc* function `icd_sums()` that tries to id rows that contain summary data over many cancer sites using ICD codes in "Sites" strings.
```{r}
## boulder has adhoc function that tries to id parent Sites
parent_nodes <- boulder::icd_sums(unique(incidence_long$Site))
parent_nodes
```

Create vector for subsetting:
```{r}
unique_parents <- unique(parent_nodes$Parent)
```

> This can be also achieved by "manually" specifying Site values to be omitted.

## Find most frequent cancer sites
We calculate incidence for each Site per Sex, let's drop Year by taking its average:
```{r}
incidence_sex_filt <- incidence_year_sex %>% 
  filter(!Site %in% unique_parents) %>% 
  group_by(Site, Sex) %>% 
  summarise(incidence = mean(incidence))
```

List of top10 cancers by Sex:
```{r}
## list of top10 cancers
listof_top10 <- incidence_sex_filt %>% 
  group_by(Sex) %>% 
  top_n(10, incidence) %>% 
  arrange(desc(incidence))
listof_top10
```

Plot most frequent cancers:
```{r}
incidence_sex_filt %>% 
  filter(Site %in% listof_top10$Site) %>% 
  ggplot(aes(Sex, incidence)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Site, labeller = label_wrap_gen()) +
  labs(title = "Top10 cancer sites from both genders",
       subtitle = "Mean annual incidence from years 2000 to 2014",
       y = "Cancer incidence per 100000",
       caption = paste("Source:", data_source)) +
  theme(axis.title.x = element_blank())
```

Here is another take using the same data:
```{r}
incidence_sex_filt %>%
  filter(Site %in% listof_top10$Site) %>% 
  ggplot(aes(reorder(Site, incidence), incidence, fill = Sex)) +
  geom_bar(stat = "identity") +
  labs(y = "Cancer incidence per 100000",
       title = "Top10 cancer sites from both genders",
       subtitle = "Mean annual incidence from years 2000 to 2014",
       caption = paste("Data:", data_source)) +
  coord_flip() +
  theme(axis.title.y = element_blank(),
        legend.title = element_blank()) +
  scale_fill_manual(values = c("darkgray", "orange"))
```

What these two graphs tell us?


## Age groups

First convert age groups to numerics:
```{r}
incidence_long <- mutate_at(incidence_long, "age", ~ parse_number(str_extract(.x, "^[0-9]{1,2}")))
```



```{r}
library(viridis)
incidence_long %>% 
  filter(Site %in% listof_top10$Site) %>%
  mutate(incidence = cases/(ave_pop/1E5)) %>% 
  group_by(Sex, Site, age) %>% 
  summarise(incidence = mean(incidence)) %>% 
  ggplot(aes(age, incidence, color = Sex)) +
  geom_line() +
  facet_wrap(~ Site, labeller = label_wrap_gen(), scales = "free_y") +
  scale_color_viridis(discrete = T)
```


Let's do the above graph using weighted incidence values.
```{r}
## calculate incidence
incidence_weighted_sex_age <- incidence_long %>%
  mutate(incidence = cases / ave_pop) %>% 
  group_by(Site, Sex, age) %>% 
  mutate(wgts = ave_pop/sum(ave_pop)) %>% 
  mutate(incidence_wgt = incidence * wgts) %>% 
  summarise(incidence_wgt = sum(incidence_wgt))
```

```{r}
## plot age group weighted incidence values
incidence_weighted_sex_age %>% 
  filter(Site %in% listof_top10$Site) %>%
  ggplot(aes(age, incidence_wgt, color = Sex)) +
  geom_line() +
  facet_wrap(~ Site, labeller = label_wrap_gen(), scales = "free_y") +
  scale_color_viridis(discrete = T)
```


