---
title: "EDA demo"
author: "Taavi Päll"
date: "27 9 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(stringr)
```


## Download dataset
```{r}
if(!file.exists("data/cacancer_newcases_PK10.json")){
  url <- "https://raw.githubusercontent.com/rstats-tartu/datasets/master/cancer_newcases_PK10.json"
  dir.create("data")
  download.file(url, "data/incidence.json")
}
```

Install boulder library to import json.
```{r}
# install.packages("devtools")
# devtools::install_github("tpall/boulder")
```

## Import dataset
```{r}
library(boulder)
incidence <- json_to_df("data/incidence.json")
```

Load tidyverse libraries:
```{r}
library(tidyverse)
library(stringi)
library(stringr) # this is what you are going to need most in future
```


For better overview let's filter out "Specified site" values that are longer than 30 characters.
```{r}
filter(incidence, str_length(`Specified site`) < 30)
```

Convert Year to numeric
```{r}
incidence <- incidence %>% mutate_at("Year", parse_number)
```

Specified site and sex to characters
```{r}
?mutate_at
incidence <- incidence %>% mutate_at(c("Specified site","Sex"), parse_character)
incidence
```

We move variables "label", "source", and "updated" to separate object for later use in plot annotation.
```{r}
label <- unique(incidence$label)
data_source <- unique(incidence$source)
updated <- unique(incidence$updated)
label
data_source
updated
```

```{r}
incidence <- incidence %>% 
  mutate(Site = stri_trim_left(`Specified site`, "[[:alpha:]]")) %>% 
  select(Year, Site, Sex, everything(), -`Specified site`, -label, -source, -updated)
incidence
```

It's not what we want...
```{r}
gather(incidence)
```

Convert incidence from wide to long: 
```{r}
incidence_long <- gather(incidence, 
                         key = age, 
                         value = cases, 
                         matches("[[:digit:]]"))
# gather(incidence, key = age, value = cases, -Year, -Sex, -Site)
incidence_long
```

Let's filter out NA-s.
```{r}
incidence_long <- drop_na(incidence_long)
```

That's all with data wrangling for now!

## Cancer trend in Estonia

Plot number of all cancer cases (Men + Women) per Year:
```{r}
total_cases_per_year <- incidence_long %>% 
  filter(Site == "All malignant neoplasms (C00-C97)") %>% 
  group_by(Year) %>% 
  summarise(total_cases = sum(cases))
total_cases_per_year
```

```{r}
p <- total_cases_per_year %>% 
  ggplot(aes(Year, total_cases)) +
  geom_point() +
  geom_line() +
  labs(title = "All malignant neoplasms (C00-C97)",
       caption = paste("Source:", data_source),
       y = label) +
  expand_limits(y = 0) +
  geom_smooth(method = "lm", se = FALSE)
p
```
Ok, trend is increasing, but how much?

Percent increase of cancer cases from 2000 to 2014? 
```{r}
cases_range <- range(total_cases_per_year$total_cases)
(1 - cases_range[1] / cases_range[2]) * 100
```

Please compare number of cases in male and females:
```{r}
total_cases_per_year_sex <- incidence_long %>% 
  filter(Site == "All malignant neoplasms (C00-C97)") %>% 
  group_by(Year, Sex) %>% 
  summarise(total_cases = sum(cases))
```

Here we replace data and aes in previous plot called "p":
```{r}
p %+% total_cases_per_year_sex + aes(Year, total_cases, color = Sex)
```

Perhaps we want to completely rewrite our plot and throw out linear model lines and expand limits option:
```{r}
total_cases_per_year_sex %>% 
  ggplot(aes(Year, total_cases, color = Sex)) +
  geom_point() +
  geom_line() +
  labs(title = "All malignant neoplasms (C00-C97)",
       caption = paste("Source:", data_source),
       y = label)
```

The __number of cancer cases__ is similar in men and women.
What about cancer incidence -- population size normalised morbidity. 
Cancer incidence is expressed as number of cases per 100000 inhabitants.
To calculate incidence, we need population size data.

Incidence is calculated like this.
```
incidence = (number of cases / mean annual population) * 100000
```

## NIH Cancer Incidence Rates
A cancer incidence rate is the number of new cancers of a specific site/type occurring in a specified population during a year, usually expressed as the number of cancers per 100,000 population at risk. That is,

Incidence rate = (New cancers / Population) × 100,000

The numerator of the incidence rate is the number of new cancers; the denominator is the size of the population. The number of new cancers may include multiple primary cancers occurring in one patient. The primary site reported is the site of origin and not the metastatic site. In general, the incidence rate would not include recurrences. The population used depends on the rate to be calculated. For cancer sites that occur in only one sex, the sex-specific population (e.g., females for cervical cancer) is used.

An age-adjusted rate is a weighted average of the age-specific rates, where the weights are the proportions of persons in the corresponding age groups of a standard population. The potential confounding effect 


Let's get population size data for year, sex and per age:
```{r}
if(!file.exists("data/mean_annual_population.csv")){
  download.file("https://raw.githubusercontent.com/rstats-tartu/datasets/master/mean_annual_population.csv", "data/mean_annual_population.csv")
}
```

Import population dataset to R:
```{r}
pop <- read_csv("data/mean_annual_population.csv")
pop
```
Population data is available from 1970 to 2016.

Summarise age groups 0 and 1-4 into age group 0-4; convert table to long format.  
```{r}
pop_long <- pop %>%
  mutate("0-4" = `0` + `1-4`) %>% 
  select(-`0`, -`1-4`) %>% 
  gather("age", "ave_pop", matches("[0-9]"))
pop_long
```

Merge cancer cases with population size data. Note that population values only from years 2000 to 2014 will be merged to incidence_long.
```{r}
incidence_long <- left_join(incidence_long, pop_long)
```

This is how you calculate incidence:
Incidence rate = (New cancers / Population) × 100,000

Cancer incidence is given as the number of new cases per 100'000.
```{r}
incidence_year_sex <- incidence_long %>% 
  group_by(Site, Year, Sex) %>% 
  summarise(cases = sum(cases),
            ave_pop = sum(ave_pop)) %>% 
  mutate(incidence = (cases / ave_pop) * 100000)
incidence_year_sex
```

All cancer cases:
```{r}
incidence_year_sex %>% filter(Site == "All malignant neoplasms (C00-C97)")
```

Cancer incidence in men versus women:
```{r}
incidence_year_sex %>% 
  filter(Site == "All malignant neoplasms (C00-C97)") %>% 
  ggplot(aes(Year, incidence, color = Sex)) +
  geom_point() +
  geom_line() +
  labs(title = "All malignant neoplasms (C00-C97)",
          caption = paste("Source:", data_source),
          y = "Cancer incidence per 100000")
```

## What are the most frequent cancer sites

We have 73 cancer sites in our dataset:
```{r}
incidence_year_sex$Site %>% unique() %>% length()
```

## What are the most frequent cancer sites?

Let's have a look at Top10 most frequent cancers.
Most frequent cancers are different in men and women.

First, we try to remove some summary variables from table.
"boulder" package has *adhoc* function `icd_sums()` that tries to id rows that contain summary data over many cancer sites using ICD codes in "Sites" strings.
```{r}
## boulder has adhoc function that tries to id parent Sites
parent_nodes <- boulder::icd_sums(unique(incidence_long$Site))
parent_nodes
```

Create vector for subsetting:
```{r}
unique_parents <- unique(parent_nodes$Parent)
```

> This can be also achieved by "manually" specifying Site values to be omitted.

## Find most frequent cancer sites
We calculate incidence for each Site per Sex, let's drop Year by taking its average:
```{r}
incidence_sex_filt <- incidence_year_sex %>% 
  filter(!Site %in% unique_parents) %>% 
  group_by(Site, Sex) %>% 
  summarise(incidence = mean(incidence))
incidence_sex_filt
```

List of top10 cancers by Sex:
```{r}
## list of top10 cancers
listof_top10 <- incidence_sex_filt %>% 
  group_by(Sex) %>% 
  top_n(10, incidence) %>% 
  arrange(desc(incidence))
listof_top10
```

Plot most frequent cancers:
```{r}
incidence_sex_filt %>% 
  filter(Site %in% listof_top10$Site) %>% 
  ggplot(aes(Sex, incidence)) +
  geom_bar(stat = "identity") + ## use the values that we have in our table
  facet_wrap(~ Site, labeller = label_wrap_gen(width = 20)) +
  labs(title = "Top10 cancer sites from both genders",
       subtitle = "Mean annual incidence from years 2000 to 2014",
       y = "Cancer incidence per 100000",
       caption = paste("Source:", data_source)) +
  theme(axis.title.x = element_blank())
```

Here is another take using the same data:
```{r}
incidence_sex_filt %>%
  filter(Site %in% listof_top10$Site) %>% 
  ggplot(aes(reorder(Site, incidence), incidence, fill = Sex)) +
  geom_bar(stat = "identity") +
  labs(y = "Cancer incidence per 100000",
       title = "Top10 cancer sites from both genders",
       subtitle = "Mean annual incidence from years 2000 to 2014",
       caption = paste("Data:", data_source)) +
  coord_flip() +
  theme(axis.title.y = element_blank(),
        legend.title = element_blank()) +
  scale_fill_manual(values = c("darkgray", "orange"))
```

What these two graphs tell us?

## Age groups

First convert age groups to numerics:
```{r}
# str_extract(unique(incidence_long$age), "^[0-9]{1,2}") %>% parse_number()
# function(x) parse_number(str_extract(x, "^[0-9]{1,2}"))
incidence_long <- incidence_long %>% 
  mutate_at("age", ~ parse_number(str_extract(.x, "^[0-9]{1,2}")))
incidence_long
```



```{r}
p <- incidence_long %>% 
  filter(Site %in% listof_top10$Site) %>%
  mutate(incidence = cases/(ave_pop/1E5)) %>% 
  group_by(Sex, Site, age) %>% 
  summarise(incidence = mean(incidence)) %>% 
  ggplot(aes(age, incidence, color = Sex)) +
  geom_line() +
  facet_wrap(~ Site, labeller = label_wrap_gen(20), scales = "free_y") +
  scale_color_manual(values = c("darkgray", "orange"))
p
```

Cancer incidence increases with age: use log on x and y axis!
```{r}
incidence_long %>% 
  filter(Site %in% listof_top10$Site) %>%
  mutate(incidence = cases / (ave_pop/1E5)) %>% 
  group_by(Sex, Site, age) %>% 
  summarise(incidence = mean(incidence)) %>% 
  mutate_at(vars(age, incidence), log) %>% 
  filter(is.finite(incidence), is.finite(age)) %>% 
  ggplot(aes(age, incidence, color = Sex)) +
  geom_point() +
  facet_wrap(~ Site, labeller = label_wrap_gen(20), scales = "free_y") +
  scale_color_manual(values = c("darkgray", "orange")) +
  geom_smooth(method = "lm")
```

There are cancer that show linear relationship between age and incidence, like skin cancer and bladder cancer.
And there are cancers that initially show increase but then reach plateau, these seem to be mostly gender specific cancers like breast cancer, cervical cancer, and endometrial cancer in women and prostate cancer in men.
If you compare breast cancer in women and men, then it seems that they have different mechanism...

## Age weighted incidence

Size distribution of different age groups in estonian population
```{r}
## install.packages("viridis")
library(viridis)
## Plot age group differences
incidence_long %>% 
  select(Year, age, Sex, ave_pop) %>%
  distinct() %>% 
  ggplot(aes(age, ave_pop, color = Sex, group = Sex)) +
  geom_jitter(jitter = 0.1) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(title = "Population sizes of different age groups in Estonia",
       subtitle = "Years 2000 to 2014",
       caption = "Data: Statistics Estonia.",
       y = "Mean annual population",
       x = "Age groups") +
  scale_color_viridis(discrete = TRUE)
```

Let's calculate age group size weighted incidence values.
```{r}
## raw incidence
raw_incidence <- incidence_long %>% 
  mutate(incidence = (cases / ave_pop) * 100000) %>% ## incidence
  group_by(Year, Site, Sex) %>% ## calculate mean population size and mean incidence
  summarise(incidence = mean(incidence),
            ave_pop = sum(ave_pop)) %>% 
  group_by(Site, Sex) %>% 
  summarise(incidence = mean(incidence),
            ave_pop = mean(ave_pop))

## Age weighted incidence
age_weighted_incidence <- incidence_long %>% 
  mutate(incidence = (cases / ave_pop) * 100000) %>% ## incidence
  group_by(Site, Sex, age) %>% ## calculate mean population size and mean incidence
  summarise(incidence = mean(incidence),
            ave_pop = mean(ave_pop)) %>% 
  mutate(wt = ave_pop / sum(ave_pop),
         incidence = wt * incidence) %>% 
  summarise(incidence = sum(incidence),
            ave_pop = sum(ave_pop))
```

```{r}
## plot age group weighted incidence values
p <- raw_incidence %>% 
  filter(Site %in% listof_top10$Site) %>%
  ggplot(aes(Sex, incidence)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Site, labeller = label_wrap_gen(), scales = "free_y") +
  scale_color_viridis(discrete = T)
p
```


```{r}
p %+% filter(age_weighted_incidence, Site %in% listof_top10$Site)
```

```{r}
bind_rows(list(raw_incidence, age_weighted_incidence), .id = "type") %>% 
  filter(Site %in% listof_top10$Site) %>%
  ggplot(aes(Sex, incidence, fill = type)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ Site, labeller = label_wrap_gen(20), scales = "free_y") +
  scale_fill_viridis(discrete = TRUE, labels = c("Raw", "Age-\nweighted"), "Incidence") +
  labs(y = "Cancer incidence per 100,000",
       title = "Top10 cancer sites from both genders",
       subtitle = "Mean annual incidence from years 2000 to 2014",
       caption = paste("Data:", data_source))
```

